---
title: "Practice - Building Blocks"
output: html_document
date: "2025-02-26"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, message = F, warning = F)
library(tidyverse)
```

**Learning assessment:** Write a named code chunk that creates a dataframe comprised of: a numeric variable containing a random sample of size 500 from a normal variable with mean 1; a logical vector indicating whether each sampled value is greater than zero; and a numeric vector containing the absolute value of each element. Then, produce a histogram of the absolute value variable just created. Add an inline summary giving the median value rounded to two decimal places. 

```{r}
df_samp = tibble (
samp = rnorm(500, mean = 1),
samp_vector = samp > 0,
numeric_samp = abs(samp_vector)
)

#histogram of the absolute  value - numeric_samp
ggplot(df_samp, aes(x=numeric_samp)) + geom_histogram()

#inline summary
median_samp = median(pull(df_samp, samp))
```

The median of the variable containing absolute value is `r round(median_samp, digits = 2)`.

**Learning assessment:** After the previous code chunk, write a bullet list given the mean, median, and standard deviation of the original random sample.

* The mean of the random sample is: `r mean(pull(df_samp, samp))`.
* The median of the random sample is: `r median(pull(df_samp, samp))`.
* The standard deviation of the random sample is: `r sd(pull(df_samp, samp))`.

